Title TBD

### 2019-07-05 main/draft.tex
Added numbers

%[Paragraph 1: is shotgun marriage still a thing?]
Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}. And, although it seems that abortion and contraception access would provide people with more choice, in the data such couples still look like something pushed them to marry each other. In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of these difference is not just a household behavior question. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 5: empirical findings --- main difference and subpopulations]
I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics. One important drawback of it is that its marriage history is not reported perfectly, as it only has records of the year of the most recent marriage. I overcome this by conditioning on women who was married exactly once. I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates. Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.

My main empirical contribution is describing association between having shotgun marriages and having a divorce. Namely, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M). I refer to the latter group as those who experienced a shotgun marriage. I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than child's father.

%Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.

The empirical result is substantially larger share of divorce people in K$\to$M group. Although this contrast is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$). This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school. When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.  %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.

%[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
Investigating the data further leads to few additional insights. First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference. Second, the divorces are concentrated around the time the child starts going to school. Third, for the cases when we can recover whether the child is step or own, this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I observe seem to have step-children as I exclude those women who had their births long ago after their marriage. [SIPP multipartner fertility data confirm these findings?]
%As additional findings, I establish few other results. [... ... ...]

@@

The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality, the importance and even presence of this channel, however, depends on parameters of individual preferences and childcare costs that I estimate structurally.

%[Paragraph 8: model --- what I get from it, mechanics and anatomy]
The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.

%[Paragraph 9: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers few nontrivial counterfactual predictions. First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

%[Paragraph 10: what I learned qualitatively]
As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes. Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children, and, although this suggestion is model-based, many things in the actual data are consistent with it.

%[Paragraph 11: what I learned about channels]
The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.

%[Paragraph 12: not a paragraph, just table of contents]
I elaborate on these points in the following order. Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6. Section 7 describes estimation results and counterfactual simulations implied by these results. Section 8 provides more general discussion and concludes.

@@

\begin{table}[h!!]
\begin{tabular}{|c|p{0.25\linewidth}|c|p{0.25\linewidth}|}\hline
Parameters & Description & Values & Identifying variation \\\hline
($\sigma_{\psi,0},\sigma_{\psi}$) & { \footnotesize
 Marriage surplus initial distribution \& shocks, standard deviations} &$(0.76,1.73)$ & {\footnotesize
 \% of divorced at $(25,35,40)$, \% of women with one marriage at 40} \\\hline\hline
$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & 
 $(0.13,0.10,0.07,0.12)$ & {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & $(-1.89,2.58)$ & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & $(0.44,1.07,-1.02)$  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  $(0.35,0.01)$ & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
$\bar{q}$ & {\footnotesize Childcare quality requirement } & $2.12$ & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & $(0.10,-3.34)$ & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
$(p^{\text{preg}}_0,p^{\text{preg}}_1,p^{\text{preg}}_2,p^{\text{preg}}_3)$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & $(0.67,0.13,-0.50,-0.005)$ & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
\end{tabular}
\caption{Estimated parameters and relevant variation.}
\end{table}

>>>>

%[Paragraph 1: is shotgun marriage still a thing? Abstract - 2]
Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage. And, although it seems that abortion and contraception access would provide people with more choice, in this paper I show that in the data such couples still look like something pushed them to marry each other. In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces. This paper have two main goals. First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups. Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of variation in divorce rates is not just a household behavior question, for instance, the divorces have large major impact on how people raise their children. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 5: empirics --- what data do I use]
I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics. One important drawback of it is that its marriage history is not reported perfectly, as it only has records of the year of the most recent marriage. I overcome this by conditioning on women who was married exactly once. I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates. Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.

%[Paragraph 6: empirics --- methodological issues shortly]
My main empirical contribution is describing association between having shotgun marriages and having a divorce, that I have not found to be explicitly discussed in the existing literature. Methodologically, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M). I refer to the latter group as those who experienced a shotgun marriage. I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than child's father.

%[Paragraph 7: empirica, findings --- main results]
The empirical result is substantially larger share of divorce people in K$\to$M group. Although this contrast is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$). This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school. When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.  %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.

%[Paragraph 8: empirical findings --- duration, step-children, lf and child development outcomes]
Investigating the data further leads to few additional insights. First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference. Second, the divorces we observe are concentrated around the time the child starts going to school. Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage. [SIPP multipartner fertility data confirms these findings?]

@@

%[Paragraph 8: model --- what generates the result in theory]
The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality, the importance and even presence of this channel, however, depends on parameters of individual preferences and childcare costs that I estimate structurally.

%[Paragraph 9: model --- what I get from it, mechanics and anatomy]
The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.

%[Paragraph 10: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers few nontrivial counterfactual predictions exploiting interaction between unplanned pregnancies and marriage quality. First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

%[Paragraph 11: what I learned qualitatively]
As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a fairly good proxy for an inferior marriage quality, and therefore the difference in divorces between two groups is not purely explained by demography and economic outcomes. Therefore, observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children. Although this suggestion is model-based, many features of the actual data are consistent with it. This does not completely rule out preference and social arguments about different attitudes to divorce and childbearing, but suggests that large part of what we see can have solid economic reasoning, and, in particular, can respond on economic policy.

%[Paragraph 12: what I learned about channels]
The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.

%[Paragraph 13: not a paragraph, just table of contents]
I elaborate on these points in the following order. Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6. Section 7 describes estimation results and counterfactual simulations implied by these results. Section 8 provides more general discussion and concludes.

@@

\begin{table}[h!!]
\begin{tabular}{|c|p{0.25\linewidth}|c|p{0.25\linewidth}|}\hline
Parameters & Description & Values & Identifying variation \\\hline
($\sigma_{\psi,0},\sigma_{\psi}$) & { \footnotesize
 Marriage surplus initial distribution \& shocks, standard deviations} &{\small$(0.76,1.73)$ }& {\footnotesize
 \% of divorced at $(25,35,40)$, \% of women with one marriage at 40} \\\hline\hline
$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & {\small
 $(0.13,0.10,0.07,0.12)$ }& {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & {\small$(-1.89,2.58)$} & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & {\small$(0.44,1.07,-1.02)$}  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  {\small$(0.35,0.01)$} & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
$\bar{q}$ & {\footnotesize Childcare quality requirement } & {\small$2.12$} & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & {\small$(0.10,-3.34)$} & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
$p^{\text{preg}}_{0,1,2,3}$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & {\small$(0.67,0.13,-0.50,-0.005)$} & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
\end{tabular}
\caption{Estimated parameters and relevant variation.}
\end{table}
